---
title: "Bias adjustment"
author: "Dan Weinberger"
date: "September 12, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Evaluation of bias adjustment vs other approach
In this set of simulations, I generate a dataset where the mean number of cases drops prior to vaccine introduction, and then the vaccine effect kicks in. I test a simple trend adjustment model, a 2 stage approach in which we first do simple trend adjustment, then fit a GAM spline to smooth the RR estimates, with or without bias adjustment. And third, a standard Interrupted time series model (change in mean and trend, with a 12 month ramp up period)


Simulate data from a Poisson model. Mean is **100 for first 90 months**, then **70 for 10 months**, then vaccine introduced, and **decline 30%** over remaining months

```{r}
set.seed(123)
post.eff.rr=seq.int(from=1, to=0.6, length.out=48 )
sim1<- c(rpois(n=90, lambda=100), rpois(n=10, lambda=70), rpois(n=48, lambda=70*post.eff.rr))
plot(sim1, ylim=c(0,max(sim1)), bty='l', bty='l')
abline(v=100, lty=2)
title('Simulated data')
```

##Method 1: fit trend to pre period; extrapolate trend, and calculate pointwise RR

```{r}
sim1.pre<-sim1
sim1.pre[101:length(sim1)]<-NA
time<-1:length(sim1)
ds1<-cbind.data.frame(sim1.pre,sim1, time)
mod1<-glm(sim1.pre~time, family='poisson', data=ds1)
mod1.pred<-exp(predict(mod1, newdata=ds1))
mod1.rr<-sim1/mod1.pred
# plot(mod1.rr)  #ESTIMATE DECLINE
# points(c(rep(1,times=100),post.eff.rr), type='l', col='#1b9e77') ##TRUE decline
# abline(h=1, v=100, col='gray', lty=2)
# abline(h=0.6, col='gray', lty=2)
```


##METHOD 2: bias correction approach in 2 stages
Take stage 1 estimate, fit spline through it

```{r}
library(mgcv)
mod1.rr.post<-mod1.rr[100:length(sim1)] #include last pre-vax time point
ds2<-cbind.data.frame(log(mod1.rr.post), (1:length(mod1.rr.post)-1))
names(ds2)<-c('log.rr','time')
mod2<-gam( log.rr~s(time), data=ds2)
pred.mod2<-exp(predict(mod2))
intercept.mod2<- log(pred.mod2[1])
pred.mod2.unbias<-exp(predict(mod2) - intercept.mod2 )
# plot(mod1.rr)  #ESTIMATE DECLINE
# points(c(rep(1,times=100),post.eff.rr), type='l', col='#1b9e77') ##TRUE decline
# points(c(rep(1,times=100),pred.mod2), type='l', col='blue') ##MODEL ESTIMATED decline
# points(c(rep(1,times=100),pred.mod2.unbias), type='l', col='#d95f02') ##MODEL ESTIMATED decline--with interept set to 1
# abline(h=1, v=100, col='gray', lty=2)
# abline(h=0.6, col='gray', lty=2)
```


#METHOD 3: ITS
```{r}
post.period<-rep(0, length(sim1))
post.period[101:112]<-1
post.period[113:length(sim1)]<-2
post.period<-as.factor(post.period)
time.all<-1:length(sim1)
ds3<-cbind.data.frame(sim1, post.period, time.all)
mod3<-glm(sim1~ post.period*time.all, family='poisson')
mod3.fit<-exp(predict(mod3, newdata=ds3))
# plot(sim1)
# points(mod3.fit, type='l', col='blue')
coef.mod3<-coef(mod3)
counter.fact<- exp(coef.mod3['(Intercept)'] + time.all*coef.mod3['time.all'] )
rr.mod3<- mod3.fit/ counter.fact
# plot(rr.mod3, type='l', col='#d95f02')  #stimate decline
# points(c(rep(1,times=100),post.eff.rr), type='l', col='#1b9e77') ##TRUE decline

```


##These plots show the True RR (blue) and the Estimated RR (orange).

```{r,fig.height = 5, fig.width = 12, echo=FALSE}
#Compare the 3 methods 

par(mfrow=c(1,4))
plot(mod1.rr, ylim=c(0.5,1.4),bty='l', type='l', col='#d95f02', lwd=2)  #ESTIMATE DECLINE
points(c(rep(1,times=100),post.eff.rr), type='l', col='#1b9e77', bty='l', lwd=2) ##TRUE decline
abline(h=1, v=100, col='gray', lty=2)
abline(h=0.6, col='gray', lty=2)
title("Linear trend extrapolation")

plot(mod1.rr, col='white',bty='l')  #ESTIMATE DECLINE
points(c(rep(1,times=100),post.eff.rr), type='l', col='#1b9e77',  ylim=c(0.5,1.4), lwd=2) ##TRUE decline
points(c(rep(1,times=100),pred.mod2), type='l', col='#d95f02', lty=1, lwd=2) ##MODEL ESTIMATED decline
abline(h=1, v=100, col='gray', lty=2)
abline(h=0.6, col='gray', lty=2)
title("2 stage: linear trend extrapolation+ GAM smooth  without bias adjustment")

plot(mod1.rr, col='white',bty='l')  #ESTIMATE DECLINE
points(c(rep(1,times=100),post.eff.rr), type='l', col='#1b9e77',  ylim=c(0.5,1.4), lwd=2) ##TRUE decline
points(c(rep(1,times=100),pred.mod2.unbias), type='l', col='#d95f02', lwd=2) ##MODEL ESTIMATED decline--with interept set to 1
abline(h=1, v=100, col='gray', lty=2)
abline(h=0.6, col='gray', lty=2)
title("2 stage: linear trend extrapolation+ GAM smooth  with bias adjustment")

plot(rr.mod3, type='l', col='#d95f02', bty='l', ylim=c(0.5,1.4), lwd=2)  #Estimate decline
points(c(rep(1,times=100),post.eff.rr), type='l', col='#1b9e77', lwd=2) ##TRUE decline
abline(h=1, v=100, col='gray', lty=2)
abline(h=0.6, col='gray', lty=2)
title("Interupted time series")
```